Contracts for unions and intersections are hard for Nickel

### 2021-06-07 paper.tex
Section 2: preamble and user-defined contracts

@@

\section{Features of languages}
\label{sec:feat-lang}
\info{Zoology of various features that we will eventually show
  conflict with this or that property or implementation of union and
  intersection. Including user-define contracts.}

Programming language research no longer lives on the abstract world
of ideal surfaces.
Every addition needs to be compatible with years of research
and development, and this is hard.
Success was unavoidable.

In this section we briefly introduce some aspects of real world
programming languages that will cause trouble while interacting
with union and intersection dynamic checks.

\subsection*{User-defined contracts}
\label{sec:flat-contracts}
\unsure{Make sure that this point is made in the user-defined
  contracts section: With respect to
validation, traditional types are too limited. Validation commonly involves
checking requirements such as being a valid port number or a valid URL which
are, while not technically impossible, really hard to check statically.}

User defined contracts, sometimes called flat contracts, are a particular
kind of contract that behaves similar to an assertion, and whose
main utility is to apply some arbitrary predicate (as arbitrary as the
host language allows it) to a value; if the predicate evaluates to
true, the contract passes, otherwise it fails.

For instance, the contract \nickel{Even} shown on Figure
\ref{fig:contract-higher-order} is a user defined contract, defined
by taking the value that needs to be checked, and validating
that is both a number and that the remainder of dividing it by 2
is 0.
And then, as seen on the same figure, we can use it to check that a function always
returns an \nickel{Int} when applied to an \nickel{Even}.

% \begin{figure}[h]
% \begin{center}
% \begin{lstlisting}[language=nickel]
% let f = fun n => n * 2 @ (Integer -> even)
% \end{lstlisting}
%   \end{center}
% \caption{A function wrapped with a contract stating that it always returns
% an even number}
% \label{fig:num-to-even}
% \end{figure}

Even more interesting, we could come up with a contract that checks that a function
maps \nickel{true} to \nickel{false} and \nickel{false} to \nickel{true}
by applying the function exhaustively and checking its results,
and then use it as a contract for the \nickel{not} function, as can be seen on Figure
\ref{fig:port-contract}.\unsure{These few paragraphs are out of date,
  even more so since Arnaud changed the flat contract example}

\begin{figure}[h]
  \begin{center}
\begin{lstlisting}[language=nickel]
let Port = contracts.fromPred (fun p =>
  num.isInt p && 0 <= p && p <= 65353) in
80 | Port
\end{lstlisting}
\end{center}
\caption{A contract for valid ports}
\label{fig:port-contract}
\end{figure}

This type of contracts are present on many different languages,
for instance, the Eiffel programming language\cite{meyer1987eiffel}, the precursor
of the Design by Contract ideals, allows to assert
these kinds of expression as pre- and post-conditions on
functions and as invariants on classes\cite{EiffelDesignByContract}.
\resolved{I (Arnaud) believe Eiffel has this feature as well,
  doesn't it? (Teo) You're right},

% A somewhat newer language, the D programming
% language, allows, among other things, to have invariants on
% classes
% written with flat contracts, that are
% dinamically checked at the frontiers of every public method \cite{DLangContracts}.

The Racket programming language also has a system to work with
contracts, powerful enough to define flat contracts
\footnote{And a gradual typing system
built on top of that, but that goes beyond the scope of this
work.}
, and
to compose them with other kinds of dynamic checks,
like higher order contracts or a lightweight take on union
and intersection contracts\cite{RacketContracts}.

Even if user defined contracts may seem like too permissive, the ability to make
non trivial checks on values (that is, to validate their output) is expected
by programmers on dynamic languages, where testing and runtime
assertions are extremely useful tools for safe programming, and
programmers are accustomed to express complex predicates
over values.

>>>>

Unfortunately, the delayed check of contract, while essential to
ensuring that schema validation doesn't affect performance (or, indeed,
is possible at all on functions), make union contracts (and their less
appreciated sibling, intersection contracts) quite problematic.

@@

\section{A typology of language features}
\label{sec:feat-lang}
\info{Zoology of various features that we will eventually show
  conflict with this or that property or implementation of union and
  intersection. Including user-define contracts.}

Union contracts are not only difficult to implement, their
unrestricted presence is incompatible with potentially desirable
properties of the language. In this section we present some of these
properties; we will show how these properties interact with union
contracts in Sections~\ref{sec:issues-sem}
and~\ref{sec:issues-literature}.

\subsection*{User-defined contracts}
\label{sec:flat-contracts}

A strength of dynamic checking is that we can easily check properties which
are impractical to check statically. For instance that a string
represents a well-formed URL, or a number is a valid port.

This same property is desirable of contracts as well, otherwise we
lose an important benefits of dynamic checking. Preferably, we want to
be able to extend the universe of contracts with user-defined
predicates (sometimes called flat contracts).

For instance, Figure~\ref{fig:port-contract} shows the definition of a
contract for valid ports in Nickel syntax.

\begin{figure}[h]
  \begin{center}
\begin{lstlisting}[language=nickel]
let Port = contracts.fromPred (fun p =>
  num.isInt p && 0 <= p && p <= 65535) in
80 | Port
\end{lstlisting}
\end{center}
\caption{A contract for valid ports}
\label{fig:port-contract}
\end{figure}

User-defined contracts can be combined with other contracts normally,
for instance \nickel{Int -> Port} is a contract verified by functions
which, given an integer returns a valid port.

This type of contracts are present on many different languages,
for instance, the Eiffel programming language\cite{meyer1987eiffel}, the precursor
of the Design by Contract ideals, allows to assert
these kinds of expression as pre- and post-conditions on
functions and as invariants on classes\cite{EiffelDesignByContract}.

The Racket programming language also has a system to work with
contracts, powerful enough to define flat contracts, and
to compose them with other kinds of dynamic checks,
like higher order contracts or a lightweight take on union
and intersection contracts\cite{RacketContracts}.
